---
title: One 'recipes' Step to Rule Them All
author: Lars Kjeldgaard
date: '2018-11-04'
slug: one-recipes-step-to-rule-them-all
categories: ["R"]
tags: ["recipes", "framework", "caret", "machine learning", "preprocessing", "modelgrid", "predictive modeling"]
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In this post I will demonstrate, how my new R package `customsteps` can be used to generate `recipes` steps, that apply custom transformations to a data set. 

Note, you should already be fairly familiar with the `recipes` package 
before reading this post or giving `customsteps` a spin!

## Introducing the 'customsteps' package

If you want to apply a transformation to your data set, that is not supported by the pre-specified steps, that come along with the `recipes` package distribution, you have two options. You could write an entire [custom `recipes` step **from
scratch**](https://github.com/tidymodels/recipes/blob/master/vignettes/Custom_Steps.Rmd), that conducts the specific transformation. This however takes quite a bit of work and code. An alternative - and sometimes better - approach is to apply the `customsteps` package, that I have just released on CRAN.

```{r, message = FALSE, warning = FALSE}
# install.packages("customsteps")
library(customsteps)
```

### Customizable Higher-Order Steps

`customsteps` contains a set of customizable higher-order `recipes` `step`
functions, that create specifications of recipe steps, that will transform or filter the data in accordance with the input functions. Let me just remind you of the definition of [**higher-order functions**](https://en.wikipedia.org/wiki/Higher-order_function):

> *In mathematics and computer science, a higher-order function is a function that does at least one of the  following: 1. takes one or more functions as arguments,
2. returns a function as its result.*

In the following I will present an example of how to use the `step_custom_transformation` function from the `customsteps` package in order to create a specification of a recipe step, that will apply a custom transformation on a data set.

## Use Case: Centering and Scaling Numeric Data

Assume, that we want to transform a variable ${\mathbf{x}}$ like this:

1. Center ${\mathbf{x}}$ around an arbitrary number $\alpha$.
2. Scale the transformed variable, such that its standard deviation equals an 
arbitrary number $\beta$.

The transformed variable $\hat{\mathbf{x}}$ can then be derived as:

$\hat{\mathbf{x}} = \alpha + (\mathbf{x} - \bar{\mathbf{x}})\frac{\beta}{s_\mathbf{x}}$

where $\bar{\mathbf{x}}$ is the mean of $\mathbf{x}$, and 
$s_\mathbf{x}$ is the standard deviation of ${\mathbf{x}}$.

**Note**, that centering ${\mathbf{x}}$ around 0 and scaling it to arrive at
a standard deviation of 1 is just a special case of the above transformation
with $\alpha = 0, \beta = 1$.

### Helper Function for the `prep` method

If the transformation requires learning parameters from the 'training data', one has to define a helper function for the `prep` method of `step_custom_transformation`. The job of the this function is - through the `prep` method - to learn the parameters, that will be used to transform a new data set. 

Obviously, the above transformation requires the mean $\bar{\mathbf{x}}$ and standard deviation $s_\mathbf{x}$ to be learned from the training data. Therefore I will write a function `compute_means_sd`, that learns the two parameters for (an arbitrary number of) numeric variables. By convention the helper function must take the argument 'x': the subset of selected variables from the training data.

```{r, message = FALSE, warning = FALSE}
library(purrr)

compute_means_sd <- function(x) {
  
  map(.x = x, ~ list(mean = mean(.x), sd = sd(.x)))

}
```
Great, we are halfway there!

### Helper Function for the `bake` method

Likewise we have to specify a helper function for the `bake` method for
`step_custom_transformation`, that defines the transformation itself and how it should be applied to a new dataset using the parameters learned from the 'training data'.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
center_scale <- function(x, prep_output, alpha, beta) {

  # extract only the relevant variables from the new data set.
  newdata <- select(x, names(prep_output))

  # apply transformation on each of these variables.
  map2(.x = newdata,
       .y = prep_output,
       ~ alpha + (.x - .y$mean) * beta / .y$sd)
  
}
```

We are all set to create a recipe drawing on the new transformation for the `mtcars` data set.

```{r, message = FALSE, warning = FALSE}
library(recipes)
rec <- recipe(mtcars) %>%
  step_custom_transformation(mpg, disp, hp,
                             prep_function = compute_means_sd,
                             bake_function = center_scale,
                             bake_options = list(alpha = 100, beta = 25),
                             bake_how = "bind_cols")
```

Prep and bake recipe.

```{r, message = FALSE, warning = FALSE}
rec_prep <- prep(rec)
rec_baked <- bake(rec_prep, mtcars)
```

Display results.

```{r}
bind_rows(
mean = map_dbl(rec_baked, mean),
sd = map_dbl(rec_baked, sd)
)
```
