---
title: One 'recipes' Step to Rule Them All
author: Lars Kjeldgaard
date: '2018-11-04'
slug: one-recipes-step-to-rule-them-all
categories: ["R"]
tags: ["recipes", "framework", "caret", "machine learning", "preprocessing", "modelgrid", "predictive modeling"]
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In this post I will demonstrate, how my new R package `customsteps` can be used to generate `recipes` steps, that apply custom transformations to a data set. 

## Introducing the 'customsteps' package

If you want to apply a transformation to your data set not included in the pre-specified steps, that come with the `recipes` package, you have two options. You could write an entire [custom `recipes` step **from
scratch**](https://github.com/tidymodels/recipes/blob/master/vignettes/Custom_Steps.Rmd) - this takes quite a bit of work and coding. Or you could just apply the `customsteps` package, that I have just released on CRAN.

```{r, message = FALSE, warning = FALSE}
# install.packages("customsteps")
library(customsteps)
```

### Customizable Higher Order Steps

`customsteps` contains a set of customizable higher order `recipes` `step`
functions, that take *helper functions* for the `recipes` `prep` and `bake`
methods as inputs and creates a customized `step` function from them. Let me
just remind you of the definition of [**higher order functions**](https://en.wikipedia.org/wiki/Higher-order_function):

> *In mathematics and computer science, a higher-order function is a function that does at least one of the  following: 1. takes one or more functions as arguments,
2. returns a function as its result.*
## An Example 

Assume, that we want to transform a variable ${\mathbf{x}}$ like this:

1. Center ${\mathbf{x}}$ around an arbitrary number $\alpha$.
2. Scale the transformed variable, such that its standard deviation equals an 
arbitrary number $\beta$.

The transformed variable $\hat{\mathbf{x}}$ can be derived as:

$\hat{\mathbf{x}} = \alpha + (\mathbf{x} - \bar{\mathbf{x}})\frac{\beta}{s_\mathbf{x}}$

where $\bar{\mathbf{x}}$ is the (estimated) mean of $\mathbf{x}$, and 
$s_\mathbf{x}$ is the (estimated) standard deviation of ${\mathbf{x}}$.

Note, that centering ${\mathbf{x}}$ around 0 and scaling it to arrive at
a standard deviation of 1 is just a special case of the above transformation
with $\alpha = 0, \beta = 1$.

Write helper function for `prep`.

```{r, message = FALSE, warning = FALSE}
library(purrr)
compute_means_sd <- function(x) {

  map(x, ~ list(mean = mean(.x),
                sd = sd(.x)))

}
```

Write helper function for `bake`.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
center_scale <- function(x, prep_output, alpha, beta) {

  newdata <- select(x, names(prep_output))

  map2(.x = newdata,
       .y = prep_output,
       ~ alpha + (.x - .y$mean) * beta / .y$sd)
  
}
```

Generate data.

```{r, message = FALSE, warning = FALSE}
df <- tibble(a = rnorm(1000),
             b = rnorm(1000),
             c = rnorm(1000))
```

Apply transformation 100, 3.

```{r, message = FALSE, warning = FALSE}
library(recipes)
rec <- recipe(df) %>%
  step_custom_transformation(everything(),
                             prep_function = compute_means_sd,
                             bake_function = center_scale,
                             bake_options = list(alpha = 100, beta = 25),
                             bake_how = "bind_cols")
```

Prep and bake recipe.

```{r, message = FALSE, warning = FALSE}
rec_prep <- prep(rec)
rec_baked <- bake(rec_prep, df)
```

Display results.

```{r}
bind_rows(
mean = map_dbl(rec_baked, mean),
sd = map_dbl(rec_baked, sd)
)
```